---
title: "Abundance Data"
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How to import abundance data?
- How to clean abundance data?
- How to summarize abundance data?
- How to vizualize abundance data?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

After following this episode, participants should be able to...

1. Import abundance data in a CSV format into R environment
2. Clean taxonomic names using the `taxize` package
    - clean spelling errors (ajr to introduce spelling errors)
    - fix synonyms 
3. Aggregate abundances
4. Calculate Hill numbers 
5. Interpret Hill numbers
6. Vizualize species abundance distributions
7. Interpret species abundance distribution plots
8. Connect species abundance patterns to Hill numbers


::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

foo bar

## Abndance data import

### Example

The example will need to show:

- how to load csv data
- different formats we might expect for abundance
- summing abundances by species and dealing with 0s
- argument options and output of `taxize::gnr_resolve`
- how to manipulate and combine different data.frames


```{r abund_load, eval = FALSE}
# load
abund <- read.csv("https://github.com/role-model/multidim-biodiv-data/raw/main/episodes/data/abundance_data.csv")

# examine
head(abund)
```


Talk about the data (fake, pretend hawaii, 4 sites, real taxa)

Talk about other data formats (site by species, 1 row per individual)

Make some pictorial examples of those data

Talk about cleaning up species names

```{r clean_taxa, eval = FALSE}
library(taxize)


wtf <- gnr_resolve(abund$GenSp)
```



Converting between data types

```{r tidy2sitespp, eval = FALSE}
# aggregation example
x <- aggregate(abund[, 'abundance', drop = FALSE], abund[, c('GenSp', 'site')], 
               sum)

# should we use tidyverse????
tidyr::pivot_wider(abund, names_from = site, values_from = abundance, 
                   values_fn = sum, values_fill = 0)

# same with base r
y <- tapply(abund$abundance, abund[, c('GenSp', 'site')], sum)
y[is.na(y)] <- 0
y[1:3, 1:3]
```

::: instructor

Here's a test of an instructor note

::::::::::::::



::: challenge

### Import and clean abundance data

Rearrange these lines of code to import and clean an abundance dataset

```{r abund-import, eval = FALSE}
names(cleanTax) <- c('sp', 'clean_name')

cleanTax <- gnr_resolve(abund$sp, 
                        best_match_only = TRUE)

cleanTax <- cleanTax[, c('user_supplied_name', 'matched_name')]


cleanTax$matched_name <- gsub(' \\(.*', '', cleanTax$matched_name)


library(taxize)

abund <- read.csv('data/abund.csv')

abund <- aggregate(list(abund = abund$n), list(sp = abund$clean_name), sum)

abund <- merge(abund, cleanTax)
```

:::::: solution

```{r abund-import-sol, eval = FALSE}
library(taxize)

abund <- read.csv('data/abund.csv')

cleanTax <- gnr_resolve(abund$sp, 
                        best_match_only = TRUE)

cleanTax$matched_name <- gsub(' \\(.*', '', cleanTax$matched_name)
cleanTax <- cleanTax[, c('user_supplied_name', 'matched_name')]
names(cleanTax) <- c('sp', 'clean_name')

abund <- merge(abund, cleanTax)

abund <- aggregate(list(abund = abund$n), list(sp = abund$clean_name), sum)
```

:::::::::::::::

:::::::::::::
