---
title: "Finding multi-dimensional biodiversity data"
teaching: 10
exercises: 2
editor_options: 
  markdown: 
    wrap: 50
---



::: questions

- Where do multi-dimensional biodiversity data live online?
- How can we combine online resources to assemble multi-dimensional datasets?
- How can we use R to retrieve data?

:::

::::::::::::::::::::::::::::::::::::: objectives

After following this episode, we intend that participants will be able to:

1. List some online repositories where different types of biodiversity data may be obtained
2. Retrieve data from GEOME & GBIF using their R APIs
3. Assemble a dataset of multiple dimensions of biodiversity from multiple online sources

::::::::::::::::::::::::::::::::::::::::::::::::


# Setup

For this episode, we'll be working with some R packages that wrap the APIs of online repositories 

```{r load packages}

library(rgbif)
library(devtools)
devtools::install_github("biocodellc/fimsR-access")
library(geomedb)

```

# Open-access data repositories

GEOME (Genomic Observatories MetaDatabase) contains genetic data associated with specimens or samples from the field. 
GBIF contains occurrance records of individuals assembled from dozens of smaller databases. 

# Downloading data using R APIs

## GBIF

Let's download some occurrence records from GBIF

## GEOME

Now let's get some genetic data from GEOME

Data in GEOME is organized into Projects, lets look at some projects

```{r geome1}

listProjects()

```

The project with ID 1 looks interesting, let's download all sequences from this project

```{r geome2}

fasta <- querySanger('CYB', projects=list(1))

```


# Assembling a combined dataset


# Recap

::: keypoints

- Many sources exist online where multidimensional biodiversity data can be obtained
- APIs allow for fast and reproducible querying and downloading
- Storing multidimensional data efficiently is important

:::
    
