---
title: "Summarizing phylogenetic data"
teaching: 10
exercises: 2
editor_options:
  markdown:
    wrap: 72
---

::: questions

- What are the common phylogeny data formats?
- How do you import and manipulate phylogeny data in R?
- How do you visualize phylogenies in R?
- How to summarize and interpret phylogenies using Hill numbers?

:::::::::::::


::: objectives

After following this episode, participants will be able to:

1. Identify key features of different phylogenetic data formats
2. Import, clean and manipulate phylogenetic data in the R environment
3. Visualize phylogenies
4. Calculate and interpret phylogenetic Hill numbers

::::::::::::::

### Introduction

In this episode, we will explore how to extract information from phylogenetic trees in order to complement our hypotheses and inferences about processes shaping biodiversity patterns. Phylogenetic trees have information about how the species in a taxonomic group are related to each other and how much relative evolutionary change has accumulated among them. Since local communities differ in their phylogenetic composition, this information can give insights on why communities are how they are.

The relative phylogenetic distance among species in a community as well as the distribution of the amount of evolutionary history (represented by the length of the branches in a phylogeny) are a result of different factors such as the age since the initial formation of the community and the rate of macroevolutionary processes such as speciation and extinction. For instance, young communities that are dominated by closely related species and show very short branch lengths may suggest a short history with few colonization events and high rates of local speciation; alternatively, if the same young communities harbor distantly related species with longer branch lengths, it may suggest that most of the local diversity was generated by speciation elsewhere, along with many colonization by species from different points in the phylogeny. Coupled with information on ecological traits and rates of macroevolutionary processes, these patterns also allow to test for hypotheses regarding, for instance, ecological filtering or niche conservatism.

> Maybe bring in the paragraph above explicit connections to our young vs older Hawaian islands examples.

Summarizing this phylogenetic information (i.e., phylogenetic distance and distribution of branch lenghts) is therefore crucial for inference. As we have seen in previous episodes, the use of [Hill numbers]() is an informative approach to summarize biodiversity. In this episode, we will then go over 1) how to import and manipulate phylogenetic data in R; 2) how to calculate and interpret Hill numbers for different communities.

### Importing phylogenetic data

Several file formats exist to store phylogenetic information. The most common formats are the [Newick]() and [Nexus]() formats. Both these formats are plain text files storing different levels of information about the taxa relationship and evolutionary history. `Newick` files are the standard for representing trees in a computer-readable form, as they can be extremely simple and therefore do not take up much memory. `Nexus` file, on the other hand, are composed by different blocks regarding different types of information and can be used to store DNA alignments, phylogenetic trees, pre-defined groupings of taxa, or everything at once. Since they are a step ahead in complexity, we will stick with `Newick` files for now.

`Newick` files store the information about the clades in a tree by representing each clade within two parentheses, with tips separated by `,`. We also add the symbol `;` at the end to represent the end of the information for that phylogenetic tree.

The basic structure of a tree in a `Newick` format is therefore as follows:

````{}
((A,B),C);
```

The notation above indicates that: 1) we have three taxa in our tree, named `A`, `B` and `C`; 2) `A` and `B` form one clade (A,B); 3) the (A,B) clade is sister to the `C` clade (we represent that by adding another set of parentheses and a `,` separating (A,B) from C).

In addition, `Newick` files can also store information on the branch leading to it tip and node. We do that by adding `:` after each tip/node.

```{}
((A:0.5,B:0.5):0.5,C:1);
```

The notation above indicates that: 1) the branches containing A and B have each a length = 0.5; 2) the branch that leads to the node connecting A and B also has length = 0.5; 3) the branch leading to C has length = 1.

The notation above is what we import into R to start working with and manipulating our phylogenetic tree. For that goal, we will use the [ape](package):

```{r loading-ape}
library(ape)
```

To import our tree, we will be using the function `read.tree()` from the `ape` package. In the case of simple trees as the one above, we could directly create them within R by giving that notation as a `character` value to this function, using the `text` argument, as shown below:

```{r read-example}
example_tree <- read.tree(text = '((A:0.5,B:0.5):0.5,C:1);')
```

Now, we can visually inspect our tree using the `plot()` function:

```{r plot-example}
plot(example_tree)
```

Can you visualize the text notation in that image? We can see the same information: A is closer related to B than C, and the branches leading to A and B have half the length of the branch leading to C.

The `read.tree()` function creates an object of class `phylo`. We can further investigate this object by calling it in our console:

```{r calling-example}
example_tree
```

The printed information shows us that we have a phylogenetic tree with 3 tips and 2 internal nodes, where the tip labels are "A, B, C". We also are informed that this tree is [rooted]() and has branch lengths.

One way to access the components of this object and better explore it is to use `$` after the object name. Here, it will be important for us to know a little bit more about where the information about tip labels and branch lengths are stored in that `phylo` objecy. Easy enough, we can access that by calling `tip.labels` and `edge.length` after `$`.

```{r calling-tips-edge}
example_tree$tip.label
example_tree$edge.length
```

Now that we learned how to import and visualize trees in R, let's bring the phylogeny for the communities we are working with in this workshop. Our data so far consists of [abundances](abundance-data.Rmd) and [traits](traits-data.Rmd) of several taxa of arthropods collected across three islands in the Hawaian archipelago. Here, we will import a global phylogeny of arthropods, located at: <>

```{r reading-phylogeny}
phylogeny <- read_tree('')
```

This new object is way larger than the previous once, being a "real" phylogeny and all. You can inspect it again by directly calling the object `phylogeny`. To plot it, we will use the `type` argument to modify how our tree will be displayed. Here, we used the option `'fan'`, to display a circular phylogeny (slightly better to show such a large phylogeny in the screen).

```{r plotting-phylogeny}
plot(phylogeny, type = 'fan')
```

How do we combine all this information with the community datasets we have so far for our three islands? First, we will have perform some cleaning.

### Cleaning phylogenetic data

Often, when we want to include the phylogenetic relatedness of organisms in our biodiversity analyses, we will run into published phylogenetic trees for our focal taxon (that is the case of the tree we are using here). A couple issues arise in these cases. First, and maybe most importantly, we need to make sure that the taxonomic names used across the different datasets in our analyses (phylogenies included) are the same. A second issue is that, depending on the resolution of the published phylogeny, it may be the case that some taxa in our [community dataset]() may not be explicitly represented in the phylogeny. In these cases, we will want to assign those "missing taxa" to one of the tips present in the tree, based on their known genus and/or family assignments.

In this section, we will deal with these issues in order to prepare our data for community inferences. Our first issue will be dealt with in a challenge!

::: challenge

In previous episodes, we have used the package `taxize` to clean and standardize the taxonomic names in out dataset. Here, we will apply the same approach to the tip names of our phylogeny. We need to make sure the names we will accept from `taxize` will match the names we have accepted in previous episodes. Finally, after we decide on the names, we need to add them to the `phylogeny` objects, where the tip names are stored.

::: solution

First, we check the names using the `gnr_resolve()` function. In `phylogeny`, tip names are stored in `phylogeny$tip.label`, so we provide that vector to our function.

```{r phylo-resolve-sol1}
library(taxize)
name_resolve <- gnr_resolve(phylogeny$tip.label, best_match_only = TRUE, canonical = TRUE)
head(name_resolve)
```

We can see what taxa are in conflict with the names retrieved by `taxize` by subsetting `name_resolve` based on boolean matching.

```{r phylo-resolve-sol2}
mismatches_phylo <- name_resolve[name_resolve$user_supplied_name != name_resolve$matched_name2, c("user_supplied_name", "matched_name2")]

mismatches_phylo
```

From here, we can see that apparently there are no misspellings in our tree (as expected from a published phylogeny). Most of the mismatches seem to be related to changes in the taxonomic treatment of species. This is a common situation, since taxonomic treatments are often revised after new trees are published; it might be the case that `nameA` in our dataset was treated as `nameB` at the time the phylogeny was created.

In the previous episodes, we used the names provided by `taxize` directly. Therefore, here we can just add those names directly to the phylogeny.

```{r phylo-resolve-sol3}
phylogeny$tip.label == name_resolve$matched_name2
```

::::::::::::

::::::::::::

The next issue we need to address is the fact that we may have species in our dataset that are not present in the phylogenetic tree we are using. We therefore need to 1) find out which species are not in the tree; 2) assign them to tips that are present in the tree.

To check which species are not in the tree, we can use any of the objects we created so far that has the list of all the species in our communities. These could be: 1) the abundances we calculated per site and island in the [abundances episode](), stored in the `abundance_tallies` object; 2) the _site by species matrix_ we have created in the abundance episode (stored in the `abundance_wide` object); 3) the list of species with their respective trait values, created in the [traits episode]() and stored in the `traits_simple` object. Since we directly use the site-by-species matrix to calculate Hill numbers (next section), let's focus on this object and retrieve the species name for it.

We do that by accessing the `colnames()` of that object. We also remove the first element, since it is just the header for the column with the `site` name.

```{r}
species_names <- colnames(abundance_wide)[-1]
```

Since the taxonomic names in both the abundance data and in the phylogenetic tree were corrected using the same database from `taxize`, we expect that any names from our abundance data that are nof found in the tree simply represent taxa that are not included in this phylogeny. We can use the boolean operator `%in%` coupled with `!` to check `species_names` that are not included in the `phylogeny$tip.label`. In summary, the expression `A %in% B` would return what elements of A are present in B. The `!` (NOT) operator returnts the opposite of that expression, in a way that `!(A %in% B)` will return the elements of A that are *NOT* present in B (which is what we want)

```{r}
not_present <- !(species_names %in% phylogeny$tip.label)
```

Now let's check those species that were not in the tree:

```{r}
species_name[not_present]
```

Here, we can decide how we are going to assign these species to tips in our tree. For instance, we see that we do not have _XXX_ in our tree, but we have _XXX_, which belongs to the same genus. Similarly...

We can then rename the species with the name of those that are present in the species.

```{r}
species_name[not_present] <- c()
```

And now we can rename the species in out site-by-species matrix using these new names. For this, we will first transfer our matrix to another object (in order to keep the abundance matrix intact), and then we will rename the columns of this new object based on the modified names.

```{r}
phylo_wide <- abundance_wide
colnames(phylo_wide) <- species_name
```

:::::::::::::::::: instructor
    
    Another issue that can arise with phylogenies is that when we assign species to a tip, we may end up with duplicates in the site-by-species matrix. We will not show this here, but it might be something important to bring up.

:::::::::::::::::::::::::::::::::::::::::::::::::::

After solving these issues, we now have 1) a phylogenetic tree with tip labels taxonomically cleaned; 2) a site-by-species matrix where all species are represented in out tree. These two objects are the two bits of information that we need to summarize the phylogenetic diversity and the distribution of branch lenghts across our focal communities.

### Summarizing with hill numbers

> Recap the idea of hill numbers. How they apply to phylogenies

As we discussed above,

"The relative phylogenetic distance among species in a community as well as the distribution of the amount of evolutionary history (represented by the length of the branches in a phylogeny) are a result of different factors such as the age since the initial formation of the community and the rate of macroevolutionary processes such as speciation and extinction. For instance, young communities that are dominated by closely related species and show very short branch lengths may suggest a short history with few colonization events and high rates of local speciation; alternatively, if the same young communities harbor distantly related species with longer branch lengths, it may suggest that most of the local diversity was generated by speciation elsewhere, along with many colonization by species from different points in the phylogeny. Coupled with information on ecological traits and rates of macroevolutionary processes, these patterns also allow to test for hypotheses regarding, for instance, ecological filtering or niche conservatism."

here we are calculating branch lengths and the distribution of those lengths across the community.

Let's first visualize what this information mean by comparing two trees with different pattern of relationships.

> Play aroud with toy trees: ladder phylogenies and perfectly balanced trees
> Totally Balanced - short and long
> Totally unbalanced
> 6 clades

Show here how to manually retrieve branch lengths and check the evenness of the distribution, as a prelude to hill numbers.

```{r}

tre1 <- read.tree(text='(a:3,(b:2,(c:1,d:1):1):1);')
tre2 <- read.tree(text='((a:1,b:1):1,(c:1,d:1):1);')


## ADD HERE CODE TO RETRIEVE BRANCH LENGHTS MANUALLY
```


> Maybe show here a SAD-like distribution of branch lengths to give an intuition of the evenness?

Values of hill numbers can give an idea of the relative length of the branches across the community and help answer the questions above.

> `hill_phylo` function - show how the function automatically calculates phylo hill numbers for each community based on a global tree. Show example in toy trees first.

```{r}
site_m <- data.frame(rbind(rep(1,4)))
colnames(site_m) <- c('a','b','c','d')

library(hillR)
hill_phylo(site_m,tre1)
hill_phylo(site_m,tre2)
```

Doing it in your own tree leaving to a challenge. Discussion on what is found?

Do the calculating for all islands in our simulated dataset using the species-by-site matrix created in previous episodes.

## Visualization of phylo Hill numbers





====
Personal code for me to better understand phylogenetic hill numbers:

```{r}
# Creating a site matrix with one site and four taxa all present in the one site
site_m <- data.frame(rbind(rep(1,4)))
colnames(site_m) <- paste0('t',seq(1:4))

# Six random trees
trees <- rmtopology(6,4,rooted = TRUE)
trees <- lapply(trees,FUN = force.ultrametric)
hill_data <- c()
for (i in 1:length(trees)) {
  hill0 <- hill_phylo(site_m, trees[[i]], q = 0)
  hill1 <- hill_phylo(site_m, trees[[i]], q = 1)
  hill2 <- hill_phylo(site_m, trees[[i]], q = 2)
  hill_data <- rbind(hill_data,c(hill0,hill1,hill2,trees[[i]]$edge.length))
}
colnames(hill_data) <- c('hill_0','hill_1','hill_2',paste0('edge_',seq(1:6)))

# Plotting SAD of branch lenghts for each tree
plot(data.frame(seq(1:6),sort(hill_data[1,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[2,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[3,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[4,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[5,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[6,4:9],decreasing = TRUE)))
```

Personal notes:
* Order 0 is just the sum of branch lengths… From order 1 and above, you weight by the abundance of that branch length in the tree… Like, how many branches have high length vs how many branchs have low length? In order one, if you have one branch that is super long and all others are super short, your hill number drops: your PD is driven by only one branch… If branch length is evenly distributed through the phylogeny (i.e., several branches with similar lenghts), then your PD increases… That is a Shannon entropy index basically…
* As you increase the order of the hill number, you are giving more weight to highly abundant branch lenghts… Order 0 is just the sum of lenghts, 1 is weight by evenness, 2 is weighted by evenness giving more weight to values of branch lengths that are more abundant, 3 is the same but focusing even more on the most abundance values (i.e., as you increase the order, you decrease your focus on rare values)…  So, like, if you have a phylogeny dminated by few branches that are very long, your hill number drops really fast as you move up in the orders… Because your hill number becomes dictated by only a few branches very fast… Whereas if you have a more even distribution of branch lenghts (several branch with similar values of length), the hill number doesn’t change as much as you move higher in the order, because even high orders still focus on most of the branches (as they are the most abundant, since they are equally abundant)
* If all branches had the exact same branch length, then hill numbers wouldn’t change as you go up in the orders (i.e, evenness is perfect)
