---
title: "Summarizing phylogenetic data"
teaching: 10
exercises: 2
editor_options: 
  markdown: 
    wrap: 72
---

::: questions 

- What are the common phylogeny data formats?
- How do you import and manipulate phylogeny data in R?
- How do you visualize phylogenies in R?
- How to summarize and interpret phylogenies using Hill numbers?

:::::::::::::


::: objectives

After following this episode, participants will be able to:

1. Identify key features of different phylogenetic data formats 
2. Import, clean and manipulate phylogenetic data in the R environment
3. Visualize phylogenies
4. Calculate and interpret phylogenetic Hill numbers

::::::::::::::

### Introduction

In this episode, we will explore how to extract information from phylogenetic trees in order to complement our hypotheses and inferences about processes shaping biodiversity patterns. Phylogenetic trees have information about how the species in a taxonomic group are related to each other and how much relative evolutionary change has accumulated among them. Such information can give insights on why communities are how they are because local communities differ in their representation of the tree of life.

Depending on different factors such as age and size of the communitu, local environmental conditions

and how we can interpret relationships of species in a community. The phylogenic history of species 



as well as charactheristics of the macoreovlutionary processes (rate of trait evolution, degree of niche conservatism, etc.)

species in a community may be closely related (phylogenetic clustering) or come from very different lineages (phylogenetic dispersion). Similarly, the amount of evolutionary history in a community (branch length)


*Phylogenetic hypotheses for community assembly*
- Phylogenetic clustering vs phylogenetic dispersion
- Amount of evolutionary history in a community may be related to age. Is it related to size?

- Do we have short branch or longer branches? That will hint into processes of community assembly.
- Is the community composed of a few long branches and several super short branches or we have several branches with similar lengths (more even distribution)? That will also hint into processes of communitu assembly.


*Hill numbers summarize shape of phylogeny*
- Recap the idea of hill numbers. How they apply to phylogenies = here we are calculating branch lengths and the distribution of those lengths across the community.
- Values of hill numbers can give an idea of the relative length of the branches across the community and help answer the questions abvoe.
- Maybe show here a SAD-like distribution of branch lengths to give an intuition of the evenness?

Do I wanna focus here on explaining everything, showing the different trees, their values, how they differ and why? Or do I show that I as we start calculating it later on (and here just show the overall concept of hill numbers)? I think the latter is better.

### File formats

Brief explanation of different phylogenetic data formats - Explain key features of each format:

* Newick
* NEXUS
* Phyllip

`ape` package

### Importing data

Reading the phylogenetic tree for our "fictional" communities

`read.tree`

```{r}
library(ape)
tree <- read.tree('tree.txt')
```

Initial plot:

`plot()`

Also show structure of `phylo` object, especially `tip.labels` (for name cleaning) and `edge.lengths` (for PD)

### Clean taxonomic names

Same code from abundance and traits episode.

:::::: challenge

`gnr_resolve()`

Save back to `tip.labels` in the same order

::: solution

:::

:::

### Pruning phylogeny to our species

`keep.tip()` based on on sites-by-spp matrix, or spp column in abundance table.

Plot again to show new tree. Show some more options to customize plot.


<<<<<<< HEAD

=======
Should I show here how to retrieve branch lengths and check the evenness of the distribution, as a prelude to hill numbers?

`ape` to do modifications
`keep.tip`
`drop.tip`
`ggtree` to visualize changes
>>>>>>> bf53cfecf6234b87348a9d489421f5d1362abcd7

### Calculating hill numbers

1. Briefly recap the meaning of hill numbers for phylogenetic trees, which was explained in intro

Should I show here how to manually retrieve branch lengths and check the evenness of the distribution, as a prelude to hill numbers? How much of that was done in previous episodes?

Question: steps 2 to 3 or jump to step 5?

2. Start with toy trees tree with 6 clades. Show `hill_phylo` on that tree. Highlight `hill_phylo` needs a spp-by-site matrix, but here we'll just make a matrix where all species of the tree are present.

<<<<<<< HEAD
5. Then move on to our simulated dataset. Highlight that there is one tree for the whole group, but different communities will have different trees. Let's check the tree for the oldest island vs the tree for the youngest island. Show how to prune the tree for the oldest island using `drop.tip` (or `keep.tip`). Challenge: do that for the other island.

6. Now plot both trees and discuss what we expect hill numbers to be. Which tree is gonna have more or less diversity?

7. Show how `hill_phylo` automatically calculates phylo hill numbers for each community based on a global tree. Do the calculating for all islands in our simulated dataset using the species-by-site matrix created in previous episodes.
=======
3. Andy's comments:
- Play aroud with toy trees: ladder phylogenies and perfectly balanced trees
Totally Balanced - short and long
Totally unbalanced
6 clades

```{r}

tre1 <- read.tree(text='(a:3,(b:2,(c:1,d:1):1):1);')
tre2 <- read.tree(text='((a:1,b:1):1,(c:1,d:1):1);')

site_m <- data.frame(rbind(rep(1,4)))
colnames(site_m) <- c('a','b','c','d')

library(hillR)
hill_phylo(site_m,tre1)
hill_phylo(site_m,tre2)

```

4. **Do this??** - Show two additional toy trees trees that has same topology as first one, but differ in branch lengths: 1) younger diversification (shorter branch lengths) and 2) older diversification (longer branch lengths). Ask them to calculate hill number for them and compare. What can we see? Here, we should be able to see branch lengths being different. How that reflect in terms of the community? Younger community vs older community?

5. Jump to `hill_phylo` function - show how the function automatically calculates phylo hill numbers for each community based on a global tree. Do the calculating for all islands in our simulated dataset using the species-by-site matrix created in previous episodes.
>>>>>>> bf53cfecf6234b87348a9d489421f5d1362abcd7


Personal code for me to better understand phylogenetic hill numbers:

```{r}
# Creating a site matrix with one site and four taxa all present in the one site
site_m <- data.frame(rbind(rep(1,4)))
colnames(site_m) <- paste0('t',seq(1:4))

# Six random trees
trees <- rmtopology(6,4,rooted = TRUE)
trees <- lapply(trees,FUN = force.ultrametric)
hill_data <- c()
for (i in 1:length(trees)) {
  hill0 <- hill_phylo(site_m, trees[[i]], q = 0)
  hill1 <- hill_phylo(site_m, trees[[i]], q = 1)
  hill2 <- hill_phylo(site_m, trees[[i]], q = 2)
  hill_data <- rbind(hill_data,c(hill0,hill1,hill2,trees[[i]]$edge.length))
}
colnames(hill_data) <- c('hill_0','hill_1','hill_2',paste0('edge_',seq(1:6)))

# Plotting SAD of branch lenghts for each tree
plot(data.frame(seq(1:6),sort(hill_data[1,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[2,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[3,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[4,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[5,4:9],decreasing = TRUE)))
plot(data.frame(seq(1:6),sort(hill_data[6,4:9],decreasing = TRUE)))
```

Personal notes:
* Order 0 is just the sum of branch lengths… From order 1 and above, you weight by the abundance of that branch length in the tree… Like, how many branches have high length vs how many branchs have low length? In order one, if you have one branch that is super long and all others are super short, your hill number drops: your PD is driven by only one branch… If branch length is evenly distributed through the phylogeny (i.e., several branches with similar lenghts), then your PD increases… That is a Shannon entropy index basically…
* As you increase the order of the hill number, you are giving more weight to highly abundant branch lenghts… Order 0 is just the sum of lenghts, 1 is weight by evenness, 2 is weighted by evenness giving more weight to values of branch lengths that are more abundant, 3 is the same but focusing even more on the most abundance values (i.e., as you increase the order, you decrease your focus on rare values)…  So, like, if you have a phylogeny dminated by few branches that are very long, your hill number drops really fast as you move up in the orders… Because your hill number becomes dictated by only a few branches very fast… Whereas if you have a more even distribution of branch lenghts (several branch with similar values of length), the hill number doesn’t change as much as you move higher in the order, because even high orders still focus on most of the branches (as they are the most abundant, since they are equally abundant)
* If all branches had the exact same branch length, then hill numbers wouldn’t change as you go up in the orders (i.e, evenness is perfect)
