---
title: "rotl name resolution"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rotl)
library(dplyr)
```

## Trying to figure out if ROTL can swap in for taxize.

```{r}

abundances <- read.csv("https://raw.githubusercontent.com/role-model/multidim-biodiv-data/rmd-review/episodes/data/abundances_raw.csv")
head(abundances)


species_list <- abundances$GenSp

short_splist <- species_list[1:10]

resolved_names <- tnrs_match_names(species_list)


```


```{r}
fuzzy_matches <- resolved_names %>%
    mutate(original_row = dplyr::row_number()) %>%
    filter((is.na(approximate_match)) | approximate_match)
```

There are 38 names without an exact match. And there are some duplicated names.

(For teaching this seems like a lot of cases to go over).

`rotl` has the `inspect` fuction for inspecting efforts to match names.


The results for `acanthia procellaris` look to me like it's matched on characters but not necessarily biologically. 


```{r}

inspect(resolved_names, row_number = fuzzy_matches$original_row[1])

```

And it didn't find subspecies...

```{r}

inspect(resolved_names, row_number = fuzzy_matches$original_row[2])


inspect(resolved_names, row_number = 173)
```

...but there's also just a lot of mismatches that, on my Googling, don't naturally match up. For example, hedylepta blackburni is a real species, and it's not immediately obvious to me why `rotl` swapped it for Adoxia femoralis. I don't think they're the same species.

For teaching, we _could_ curate a species list with only a few, teachable-moment mismatches.


****

O.M.G. https://mctavishlab.github.io/R_OpenTree_tutorials/01-find-taxa/index.html
